---
title: "Albrecht_Capstone_Final_Report"
author: "Evelyn Albrecht"
format: html
editor: visual
---

## Introduction

For this analysis, I worked with the Frugivoria trait databases. Compiling data from 1924 to 2023, Beth Gerstner, Pheobe Zarnetske and Patrick Bill sought to create a robust dataset which would allow for the quantifying of functional diversity within ecosystems. Frugivoria is a trait database, recording information on “dietary, life-history, morphological, and geographic traits” of frugivorous birds and mammals in Central and South America. Frugivoria serves as a resource to make research involving these species more accessible.

Withing Frugivoria, observations are divided between two taxa, mammals and birds. Each taxa had its own complete database and a second that narrowed down variables that were accounted for in both mammal and bird datasets.

I worked with both the mammal and bird datasets in their own separate analyses.

### Data Preparation

To prepare the data for analysis, I first omitted most variable that did not pertain to my planned analyses. This was mainly for convince as these bases both had over 50 variables. No extra alteration of the original datasets was needed.

In the data exploration stage, I determined that any "outliers" could be rationalized as natural variation and thus were not eliminated from the dataset. However, for the mammal database, I omitted observations whose IUCN status was Data Deficient because that was such a central aspect to the analysis. (the bird database did not have any species in this category so therefore that particular alteration was not necessary)

# Analysis 1

## Introduction

In this first analysis, I will be looking at the relationships between diet breadth, human footprint, and IUCN status in neotropical mammals displaying frugivory to some extent. I anticipate species with wider diet breadths and lower human footprints to rank in lower IUCN classes. Cloyed et al. argues the species with narrower diet breadths are more susceptible to anthropogenic changes whilst species with broader dietary breadths are more resistant because they have more options for resources to turn to in the case one is becomes scarce or extirpated all together. the study was on cetaceans, how does this track with terrestrial mammals in the neotropics. In order to explore this relationship, I will be using ordinal regression to plot these two continuous variables as predictors to a categorical ordinal response one.

IUCN status is an ordinal categorical variable, the ranks worsening in severity. From least to greatest extinction risk, the categories are as followed.

NT - Near threatened

VU - Vulnerable

EN - Endangered

CR - Critically Endangered

### Hypothesis:

Both diet breadth and human footprint will have an impact on the mammal's IUCN ranking.

### Null Hypothesis

Neither diet breadth nor human footprint will have an impact on the mammal's IUCN ranking.

## Setting up

Let's clear R's environment and open relevant packages.

All packages were installed in the console window prior to creating this qmd.

```{r}
rm(list=ls())
library(tidyverse)
library("here")
library(MASS)
library(Hmisc)
library(IDPmisc)
library(oddsratio)
library(visreg)
```

Read in the data and assign it a name.

```{r}
mammals <- read.csv(here("Data","Frugivoria_mammals_edited_again.csv" ))
```

Take a quick look at the data

```{r}
glimpse(mammals)
```

We have 122 observations of 13 variables. Of those 13 variables, 4 are nominal and 9 are categorical. For this analysis we are using one categorical and two numeric variables, IUCN_category, diet_breadth, and mean_human_fp_range(2020 version)

## Guess Relationships

### Create an Initial Plot

#### Transform response variable into ordered factor

For an ordinal regression, it is important to let R know that our response variable is is ordinal. Therefore the first step should be to convert it from the default order (alphabetical) to increasing likelihoods of extinction.

```{r}
mammals$IUCN_category <- factor(mammals$IUCN_category,
                                levels = c("NT","VU","EN","CR"),
                                ordered = TRUE)
```

Now we can see that IUCN_category is an ordered factor ranking in terms of risk of extinction from least to greatest.

Plot a histogram of our response variable. In this case it is IUCN status.

```{r}
ggplot(mammals, aes(IUCN_category))+
  geom_histogram(stat="count")+ 
  theme_bw()
```

We have quite the non-normal distribution here however it is not abnormal to what we would expect. Summary statistics for the IUCN red list demonstrate that the Vulnerable and Endangered categories composes the highest percentage of species (out of these four rankings). We can also note that our variable is an ordered one.

(<https://www.iucnredlist.org/resources/summary-statistics)>

In the class example, the next step would be to create a table to summarize the the frequencies in which observations fall into different categories. This will not work with my data because both of my predictors are nominal and thus R would create a new column/row for each unique number and will be counterproductive in trying to better understand the data by visual means.

Here is the formula that would theoretically be used:

**Do Not Run!**

```{r}
ftable(xtabs(~response + predictor_1 + predictor_2, data=data))
```

### Make plot of predictors and response variable

plot the predictor and response the variable one at a time because the faceting will not work with numerical variable

Once again, my predictor variables both being numerical complicates exactly following the example steps. facet_grid() which would allow for use to compare our response against a second predictor only works with categorical variables. I will instead opt to create two graphs each plotting one predictor against the response.

**Human footprint first**

```{r}
ggplot(mammals, aes(x=IUCN_category, y=mean_human_fp_range_2020)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() + 
  theme(axis.text = element_text(angle = 45, hjust = 1, vjust = 1))
```

There does not seem to be much of an observable trend. critically endangered species seemingly experince the highest human footprint. However the distribution of observations is very scattered. Vulnerable species have an unexpectedly low experience of human footprint.

**Now diet breadth**

```{r}
ggplot(mammals, aes(x=IUCN_category, y=diet_breadth)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() + 
  theme(axis.text = element_text(angle = 45, hjust = 1, vjust = 1))
```

There does not appear to be much of a pattern Vulnerable, ENdangered, and CRitically endangered having very similar means.

## Create Model

Using the proporitional odds logistic regression (aka. polr()) function, create model measuring IUCN category as a factor of human footprint and diet breadth.

```{r}
mammal_model <- polr(IUCN_category ~ mean_human_fp_range_2020 + diet_breadth,data = mammals, Hess = TRUE)
```

## Check Assumptions

Checking the assumptions of this model is far beyond the scope of this course and thus I will not attempt to check them. Just note it is at this point that I would check assumptions of the model.

## Run Model

Obtain summary statistics from our model using the summary() function.

```{r}
summary(mammal_model)
```

Here, we are given our coefficients for our predictor variables, intercepts ("cutpoints" between the ordered factors), and residual deviances.

### Calculate p-values and interpret

Lets calculate some p-values to better understand what these outputs indicate about significance. We will calculate our p-values by comparing our t-values to a normal distribution.

Create a table from the output given when we used summary() to run the model)

```{r}
ctable <- coef(summary(mammal_model))
```

Calculate confidence intervals using the liklihood method.

```{r}
confidence_intervals <- confint(mammal_model)
View(confidence_intervals)
```

This gives us two numbers, on at 2.5% and 97.5%. This interval accounts for 95% of the data. If the transition from the former to the latter includes the value 0 then it is not a significant predictor of our response variable.

Both human footprint and diet breadth includes zeros and are thus not significant predictors of IUCN category.

This does not allow use to reject our null hypothesis.

At this point, we would rebuild the model with only significant predictors and obtain the p-values from that model and then move onto building a final plot. Since there are no significant predictors, I will not move forward.

A possible rationale for the data not supporting is the relatively small sample size. I started out with only 122 observations and once 4 were taken away due to missing values while running the model I was down to 118.

# Analysis 2

## Introduction

In this analysis, I wish to test the effect of diet breadth and geographic range in neotropical birds on a this set to analyze the relationship between diet breadth and geographic range. This analysis builds off of the breadth range theory that suggests that animals with larger diet breadths are likely to have a larger geographic range because they are able to survive in a broader array of places offering diverse diets.

### Hypothesis

I predict that the data will demonstrate a positive relationship between diet breadth and geographic range. I expect to see that species with broader diet breadths also will have larger geographic ranges.

### Null Hypothesis

There is no correlation between diet breadth and geographic range in neotropical birds.

## Setting Up

Clear R's environment and open necessary packages

```{r}
rm(list = ls()) 
library(tidyverse) 
library("here")
library(ggfortify)
```

Read in the data for the analysis and give it a name.

```{r}
birds <- read.csv(here("Data","frugivoria_birds_edited.csv"))
```

## Initial Plotting

**Make an initial plot of the two variables being used in this analysis.**

Lets start by looking at a plot of the distribution of the response variable.

```{r}
ggplot(birds, aes(observed_range_sqkm)) + 
  geom_histogram() + 
  theme_bw()
```

This distribution is very abnormal. I will adress this issue once I get to the step of building my model. For now I will continue and plot my response variable again my predictor one.

```{r}
ggplot(birds, aes(x=diet_breadth, y=observed_range_sqkm))+ 
  geom_point()+
  theme_bw()
```

**Guesstimate slope of a theoretical line that would represent the trend of the data.**

I am not seeing much of a pattern right now so let's just calculate based off of the ranges of our x and y variables.

Diet breadth ranges from about 0 to about 2.05

Observed geographic range ranges from 1 to 19000000 square kilometers

lets use these ranges to calculate a *very* rough estimate of a slope.

y2 - y1 / x2 - x1

(19000000 - 1)/(2.05 - 0)

Our estimated slope is 18999999/2.05 = 9268292. This means with each increase of 1 diet breadth metric, we anticipate a species home range to increase 9268292 square kilometers.

Its very difficult to pinpoint a y-intercept but it looks to be close to zero.

Let's now alter the graph to add a both a linear and non-linear line of best fit.

```{r}
ggplot(birds, aes(x=diet_breadth, y=observed_range_sqkm))+ 
  geom_point()+
  geom_smooth(method = "lm", se=F) + 
  geom_smooth(span = 1, color = "red", se = F) + 
  xlab("Diet breadth") + 
  ylab("Observed geographic range (km2)")
```

This is not looking fantastic, neither line seems to capture a trend. I am not optimistic will keep moving through the typical workflow and move to creating the model.

## Create Model

I will initially create a plain linear model. Using the autoplot() function, I check if the data meets assumptions required to conduct a linear regression.

```{r}
birds_lm <- lm(observed_range_sqkm ~ diet_breadth, data = birds)
```

### Check Assumptions

```{r}
autoplot(birds_lm)
```

The normal Q-Q graph not following the dashed line indicates that the data does not follow a normal distribution.

**Lets craft a glm (generalized linear model) and see if we can find a model that better fits the distribution of our data.**

The poisson family model is a non-linear model that has utility when data's upper value us unbounded.

Before crafting a glm using the poisson family, get some summary statistics for our predictor variable.

```{r}
bird_summary <- birds %>% summarise(
  mean_diet_breadth = mean(diet_breadth, na.rm = T),
  sd_diet_breadth = sd(diet_breadth, na.rm = T)
)
view(bird_summary)
anova(birds_lm)
summary(birds_lm)
```

See if the Poisson model better represents th data.

```{r}
birds_glm <- glm(observed_range_sqkm ~ habitat_breadth, data = birds, family = poisson)
```

I am assaulted with a screen full of warnings about non-integers.

Attempt to auto plot again and see if our Normal Q-Q plot looks any better.

```{r}
autoplot(birds_glm)
```

The Normal Q-Q plot looks better but still not great. Additionally, the other plots are now looking more abnormal.

Lets try to run the model anyway and see what happens.

## Run Model

```{r}
anova(birds_glm)
summary(birds_glm)
```

There is alot overdispersion! Yikes! Lets calculate the residual index anyway by dividing the residual deviance by the degrees of freedom. The class text Getting Started With R (GSWR) says a residual index over two is a sign to worry.

A residual index of 3380827 is highly alarming.

## Interpret Model

Theoretically, I would move forward and interpret my model and plot my final results. that is not going to happen due to the extent to which my data is not-normal.

# Analysis 3 

## Introduction

In this third and final analysis, I will be looking at multiple factors contributing to neotropical birds extinction risk.A large body size, long amount of time between generations, small home range, and a diet containing a high prevalence of fruit are all factors that contribute to a birds susceptibility to extinction in the face of increased human disturbance(Newbold et al. 2013).

### Hypothesis

Longevity, generation time, home range size and percent fruit in diet will serve as predictors of IUCN status, with higher measurements of these variables corresponding to more severe IUCN categorization. \### Null Hypothesis Longevity, generation time, home range size and percent fruit in diet will have no impact on IUCN categorization.

## Setting UP

Clear R's environment and open relevant packages.

```{r}
rm(list=ls())
library(tidyverse)
library("here")
library(MASS)
library(Hmisc)
library(IDPmisc)
library(oddsratio)
library(visreg)
```

Read in the data and assign it a name.

```{r}
birds <- read.csv(here("Data","frugivoria_birds_edited.csv"))
```

## Initial Plots

### Transform response variable into an ordered factor

Since this is another ordinal regression, I will start letting R know that our response variable is is ordinal. Therefore the first step should be to convert it from the default order (alphabetical) to increasing likelihoods of extinction.

```{r}
birds$IUCN_category <- factor(birds$IUCN_category,
                                levels = c("LC", "NT","VU","EN","CR"),
                                ordered = TRUE)
```

### Guess Relationships

Like the first analysis, all of my predictor variables are nominal and thus I will skip creating a frequency table for the same reasons stated there."This will not work with my data because both of my predictors are nominal and thus R would create a new column/row for each unique number and will be counterproductive in trying to better understand the data by visual means."

Additionally, I will also be making each plot with only one predictor variable because they are all nominal and facet_grid() does not work on nominal variables.

```{r}
ggplot(birds, aes(x=IUCN_category, y=diet_fruit_e)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
  
```

There appears to be little correlation between percent fruit in diet and IUCN status.

```{r}
ggplot(birds, aes(x=IUCN_category, y=longevity)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
```

There appears to be a slight correlation between longevity and IUCN status, longer lived species having greater extinction risks.

```{r}
ggplot(birds, aes(x=IUCN_category, y=home_range_size)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
```

Home range size does not appear to be a predictor IUCN category

```{r}
ggplot(birds, aes(x=IUCN_category, y=generation_time)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
```

It does not appear as though generation time is a predictor of IUCN status.

## Create Model

Using the proportional odds logistic regression (aka. polr()) function, create model measuring IUCN category as a function of percent diet fruit, body size, longevity, home range size, and generation time.

```{r}
bird_model <- polr(IUCN_category ~ diet_fruit_e + body_size_mm + longevity + home_range_size + generation_time, data = birds, Hess = TRUE)
```

## Check Assumptions

Checking the assumptions of a polr() model is far beyond the scope of what was taught in this class and therefore this step will be acknowledged and ommitted.

## Run Model

Obtain summary statistics from our model using the summary() function.

```{r}
summary(bird_model)
```

Here, we are given our coefficients for our predictor variables, intercepts ("cutpoints" between the ordered factors), and residual deviances.

### Calculate p-values and interpret

Lets calculate some p-values to better understand what these outputs indicate about significance. We will calculate our p-values by comparing our t-values to a normal distribution.

Create a table from the output given when we used summary() to run the model)

```{r}
ctable <- coef(summary(bird_model))
```

```{r}
confidence_intervals <- confint(bird_model)
```

Calculate confidence intervals using the liklihood method.

```{r}
confint.default(bird_model)
```

This gives us two numbers, on at 2.5% and 97.5%. This interval accounts for 95% of the data. If the transition from the former to the latter includes the value 0 then it is not a significant predictor of our response variable.

All of our predictor variables include 0 in their ranges therefore none of these variables are significant indicators of IUCN status.

I would go on to calculate odds ratios at this point but I already know there is nothing of significance.

## Interpret Model

I found that none of the variables I

## Rebuilding the model

Since so many observations were lost due to missingness, I am going to quantify NAs, I am going to see if removing any of them will do anything to possibly allow me to find significant results.

### Find NAs

Use the complete cases function to determine the amount of NAs for each variable.

```{r}
sum(is.na(bird_df$home_range_size))
```

```{r}
sum(is.na(bird_df$body_size_mm))
```

```{r}
sum(is.na(bird_df$longevity))
```

```{r}
sum(is.na(bird_df$generation_time))
```

```{r}
sum(is.na(bird_df$diet_fruit_e))
```

let re-run the model without home range and longevity because of how much they eliminate. the longevty may slightly negatively effect on icun

```{r}
bird_model2 <- polr(IUCN_category ~ diet_fruit_e + body_size_mm + generation_time, data = birds, Hess = TRUE)
```

```{r}
summary(bird_model2)
```

### I want p-values

```{r}
ctable <- coef(summary(bird_model2))
```

### Using likelihood method to find confidence intervals

```{r}
confidence_intervals <- confint(bird_model2)
```

```{r}
confint.default(bird_model2)
```

### calculate odd ratios

```{r}
exp(coef(bird_model2))
```

```{r}
exp(cbind(OR = coef(bird_model2),confidence_intervals))
```

## Interpret Model

# Conclusion

### Challenges

I encountered many challenges throughout this process. Most of them were related to either the questions I wished to ask or the nature of my data set itself. Whilst I think that ordinal regression is cool, It was very intimidating to figure out, and even then I wasn't the one to fully figure it out (thank you so much for all the help with that). That was a situation unique to me as the only one doing ordinal logistic regression.

Additionally, I think that using a trait database as opposed to a more typical dataset set me up for complications specifically regarding the normality of my data. I think that the high percentages of species ranking data deficient of least concern also limited my number of observations and made it hard to analyze. It was also very unfortunate I didn't end up getting anything significant and thus didn't get to have fun making graphs.

### Works Cited

Cloyed CS, Balmer BC, Schwacke LH, Wells RS, Berens McCabe EJ, Barleycorn AA, Allen JB, Rowles TK, Smith CR, Takeshita R, et al. 2021. Interaction between dietary and habitat niche breadth influences cetacean vulnerability to environmental disturbance. Ecosphere. 12(9):e03759. doi:10.1002/ecs2.3759. \[accessed 2024 Oct 26\]. https://onlinelibrary.wiley.com/doi/abs/10.1002/ecs2.3759.

Gerstner, B.E., P.L. Zarnetske, and P. Bills. 2023. Frugivoria: A trait database for birds and mammals exhibiting frugivory across contiguous Neotropical moist forests ver 6. Environmental Data Initiative. <https://doi.org/10.6073/pasta/66423109f351d2fdc0fdbe9c4001e19e> (Accessed 2024-10-25).

Huang S, Tucker MA, Hertel AG, Eyres A, Albrecht J. 2021. Scale-dependent effects of niche specialisation: The disconnect between individual and species ranges. Ecology Letters. 24(7):1408–1419. doi:[10.1111/ele.13759](https://doi.org/10.1111/ele.13759). \[accessed 2024 Dec 19\]. <https://onlinelibrary.wiley.com/doi/abs/10.1111/ele.13759>.

Newbold T, Scharlemann JPW, Butchart SHM, Şekercioğlu ÇH, Alkemade R, Booth H, Purves DW. 2013. Ecological traits affect the response of tropical forest bird species to land-use intensity. Proc R Soc B Biol Sci. 280(1750):20122131. doi:10.1098/rspb.2012.2131. \[accessed 2024 Oct 26\]. https://royalsocietypublishing.org/doi/full/10.1098/rspb.2012.2131.
