---
title: "Analysis_draft_analysis_three"
format: html
editor: visual
---

## Introduction

In this third and final anaylsis, I will be looking at multiple factors contributing to neotropical birds extinction risk.

```{r}
rm(list=ls())
library(tidyverse)
library("here")
library(MASS)
library(Hmisc)
library(IDPmisc)
library(oddsratio)
library(visreg)
```

```{r}
birds <- read.csv(here("Data","frugivoria_birds_edited.csv"))
```

## Initial Plot

Identify NAs before starting analysis - not using all of the data complicates

## Guess Relationships

```{r}
birds$IUCN_category <- factor(birds$IUCN_category,
                                levels = c("LC", "NT","VU","EN","CR"),
                                ordered = TRUE)
```

```{r}
ggplot(birds, aes(x=IUCN_category, y=diet_fruit_e)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
  
```

```{r}
ggplot(birds, aes(x=IUCN_category, y=diet_inv_e)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
```

```{r}
ggplot(birds, aes(x=IUCN_category, y=body_size_mm)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
```

```{r}
ggplot(birds, aes(x=IUCN_category, y=longevity)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
```

```{r}
ggplot(birds, aes(x=IUCN_category, y=home_range_size)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
```

```{r}
ggplot(birds, aes(x=IUCN_category, y=generation_time)) + 
  geom_boxplot() +
  geom_jitter() +
  theme_bw() 
```

## Create Model

```{r}
bird_model <- polr(IUCN_category ~ diet_fruit_e + body_size_mm + longevity + home_range_size + generation_time, data = birds, Hess = TRUE)
```

## Check Assumptions

Checking the assumptions of a polr model is far beyond the scope of what was tought in this class and therefore this step will be aknowldeged and ommited. 

## Run Model

```{r}
summary(bird_model)
```

### I want p-values

```{r}
ctable <- coef(summary(bird_model))
```

### Using likelihood method to find confidence intervals

```{r}
confidence_intervals <- confint(bird_model)
```

```{r}
confint.default(bird_model)
```

All of these include 0 so nothing is a significant indicator of IUCN status

### calculate odd ratios

```{r}
exp(coef(bird_model))
```

```{r}
exp(cbind(OR = coef(bird_model),confidence_intervals))
```

## Interpret Model

nothing matter. generation time has a little bit of an effect

could have something to do with sampling size 1031 rows were deleted leading to a prett small set to work with (\<150)

this isnt good. im fine though. it joever fr fr

AHHHHHH

## Final Plot

Again, but adressing the NAs cause maybe that could be the problem

## Attempt 2 

read in the data and glimpse the structure

```{r}
bird_df <- read.csv(here("Data", "frugivoria_birds_edited.csv"))
str(birds_df)
```

x rowa , x col

at this point i would convert cat to cat all numeric so i dont have to

use the complete cases function

```{r}
sum(is.na(bird_df$home_range_size))
```

927

```{r}
sum(is.na(bird_df$body_size_mm))
```

```{r}
sum(is.na(bird_df$longevity))
```

```{r}
sum(is.na(bird_df$generation_time))
```

```{r}
sum(is.na(bird_df$diet_fruit_e))
```

let re-run the model without home range and longevity because of how much they eliminate. the longevty may slightly negatively effect on icun

```{r}
bird_model2 <- polr(IUCN_category ~ diet_fruit_e + body_size_mm + generation_time, data = birds, Hess = TRUE)
```


```{r}
summary(bird_model2)
```

### I want p-values

```{r}
ctable <- coef(summary(bird_model2))
```

### Using likelihood method to find confidence intervals

```{r}
confidence_intervals <- confint(bird_model2)
```

```{r}
confint.default(bird_model2)
```
### calculate odd ratios

```{r}
exp(coef(bird_model2))
```

```{r}
exp(cbind(OR = coef(bird_model2),confidence_intervals))
```

## Interpret Model
